Incremental tactics

### 2012-05-31 doc/papers/LFMTP2012/paper.tex
Revert some debug message

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of 1/ an untrusted procedure $\finfer M$ to infer a typing
derivation out of program $M$; it doesn't have to be complete, or even
sound, and 2/ a trusted and fast kernel $\fcheck{\md}$ to verify these
derivations in a given set of inference rules $\Sigma$. This is a
\emph{certifying} scheme: %TODO cite Leroy
if $\fcheck{\finfer M} = \cst{true}$ then there really \emph{is} a
derivation $\md$ for $M$, but the converse is not necessarily true:
function $\finfer{}\!\!$ could fail, or have bugs (return an incorrect
derivation). Explicitly generating the derivation in the declarative
system has the following advantages over the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping those types. He (or
she) wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but \sysname{T_{<:}} is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's
provided incremental verifier. As a first attempt, he wants to verify
$P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \\&\qquad
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for then
new recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{even}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
system will generate this derivation, it will insert a coercion at
these two points.
% TODO j'en ss la ds la relecture F<
Taking advantage of $\md_2$ and $\md_4$ the system
succeeds with a new derivation made out of these, without having to
regenerate them because it knows that $\finfer{\fget\md} = \md$.

@@

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so we just have to
use it here.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} x y {\s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

A few days earlier, the teacher is preparing the tool for the students
to train composing System \sysname{T} programs. He (or she) provides
the abstract syntax, the typing rules and the typing untrusted
algorithm in order to build the incremental type checker. How should
he write the algorithm? We know that we want to return a derivation,
and take advantage of a higher-order representation of the syntax, so
it shouldn't be necessary to thread any explicit environment. The
definition should then start with:

@@

\noindent
It takes a term $M$ and producing the pair of a type $A$ and a
derivation that $M$ has type $A$. Its code will be untyped, but
the teacher provides this type so that the checker can verify that
each call is fed and returns values of the right type: it is its
\emph{contract} so to say. He first decomposes term $M$ and treats
the easy cases of $\cst{o}$, $\cst{s}$ and $\cst{app}$:

\begin{mathleft}
  \lamd M \match{M} \\
  \quad\case{\z} \pair {\cst{nat}}{\infer{ }{\vdash \cst o : \cst{nat}}} \\
  \quad\case{\s M}
  \letd {\pair {\cst{nat}} \md} {\finfer{M}}
  \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
  \quad\case{\app{M}N}
  \letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}}
\end{mathleft}

\noindent
We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types; the
$\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; patterns don't have to be exhaustive since we are in
an untrusted setting; for conciseness, we use case failure for
signaling typing error; we use an auxiliary function $\!\!\feq{}{}\!\!
: \cst{tp}\to\cst{tp}\to\cst{bool}$ for syntactic comparison of types,
defined the same way.

@@

The case of $\cst{rec}$ is similar: we first compute recursively the
derivations $\md_M$ and $\md_N$ which should be well-typed in the
current environment, and then $\md_P$ in the enlarged environment,
with variables $\var x$ and $\var y$ in $P$ substituted with the
(future) result of $\finfer{}$.

\begin{mathleft}
  \quad\case{\recb M N x y P} \\
  \quad\quad
  \letd {\pair {\cst{nat}} {\md_M}} {\finfer{M}} \\
  \quad\quad
  \letd {\pair {A_1} {\md_N}} {\finfer{N}} \\
  \quad\quad
  \syntax{let} {\pair {A_2} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_1)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A_1}{\md_y}}}}}} \\
  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}} \\
  \quad\quad
  \pair {A} {
    \infer{
      \md_M \and
      \md_N \and
      \mbox{$
        \begin{array}{c}
          [\md_x] [\md_y] \\
          \md_P
      \end{array}
        $}
    }{
      \vdash \recb M N x y P
    }
  }
\end{mathleft}

@@

We exploit here an idea similar to run-time
\emph{contracts}: %TODO cite
there is no static guarantee that computations won't return ill-typed
values, but arguments and results are checked at run-time at the
boundaries of defined functions. This allows to omit formal
justifications, like the fact that $B$ can't depend on $\md_x$ in the
$\lambda$ case, but of course incurs on the performance. This is
compensated by the slicing mechanism described below in
\ref{sec:repr}.

Introducing the ``inverse'' of the derivation inference function has
two purpose: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would arrive to the unhandled case $\finfer{\var
  x}$. This requires evaluating the type-checker with a carefully
designed strategy.

For the sake of conciseness, we took a simple, syntax-directed system
as example; this scales nonetheless to declaratively presented
features like subtyping or conversion.

@@

\subsection{The computational calculus}

>>>>

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of 1/ an untrusted procedure $\finfer M$ to infer a typing
derivation out of program $M$; it doesn't have to be complete, or even
sound, and 2/ a trusted and fast kernel $\fcheck{\md}$ to verify these
derivations in a given set of inference rules $\Sigma$. This is a
\emph{certifying} scheme (see \cite{leroy2006compcert} for a
discussion on approaches to reliable compilers):
if $\fcheck{\finfer M} = \cst{true}$ then there really \emph{is} a
derivation $\md$ for $M$, but the converse is not necessarily true:
function $\finfer{}\!\!$ could fail, or have bugs (return an incorrect
derivation). Explicitly generating the derivation in the declarative
system has the following advantages over the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types. He (or
she) wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but \sysname{T_{<:}} is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's
provided incremental verifier. As a first attempt, he wants to verify
$P_1$:
$$
\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \\&\qquad
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for then
new recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
system will generate this derivation, it will insert a coercion
$\cst{odd} \leq \cst{nat}$ at these two points. Taking advantage of
$\md_2$ and $\md_4$ the system succeeds with a new derivation made out
of these, without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

@@

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so we just have to
use it here. $\md_2$ was deriving $\vdash\s\z:\cst{nat}$ (with a
coercion since it was used as an argument to $\var{add}$); the system
thus infer the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one into the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} x y {\s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

A few days earlier, the teacher is preparing the tool for the students
to train composing System \sysname{T_{<:}} programs. He (or she)
provides the abstract syntax, the typing rules and the untrusted
typing algorithm in order to build the incremental type checker. How
should he write the algorithm? We know that we want to return a
derivation, and take advantage of a higher-order representation of the
syntax, so it shouldn't be necessary to thread any explicit
environment. The definition should then start with:

@@

\noindent
Function $\finfer{}$ takes a term $M$ and produces the pair of an
inferred type $A$ and a derivation that $M$ has type $A$. Its code
will be untyped, but the teacher provides this type so that the
checker can verify that each call is fed and returns values of the
right type: it is its \emph{contract} so to say. He first decomposes
term $M$ and treats the easy cases of $\cst{o}$ and $\cst{s}$:

\begin{mathleft}
  \lamd M \match{M} \\
  \quad\case{\z} \pair {\cst{even}}{\infer{ }{\vdash \cst o : \cst{even}}} \\
  \quad\case{\s M}
  \match{\finfer M} \\
  \quad\quad\case{\pair{\cst{even}}\md}
  \pair {\cst{odd}}{\infer\md{\vdash\app{\cst{s}} M : \cst{odd}}} \\
  \quad\quad\case{\pair{\cst{odd}}\md}
  \pair {\cst{even}}{\infer\md{\vdash\app{\cst{s}} M : \cst{even}}} \\
  \quad\quad\case{\pair{\cst{nat}}\md}
  \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
\end{mathleft}

\noindent
We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types;
patterns don't have to be exhaustive since we are in an untrusted
setting; for conciseness, we use case failure for signaling typing
error. There are three cases for $\s{M}$ in the system, corresponding
to the three rules in the declarative system. For the application
case, we rely on the well-known property that we may need to insert a
coercion from the actual type of the argument to its expected type:

\begin{mathleft}
  \quad\case{\app{M}N} \\
  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
  \quad\quad
  \match{\md_{\leq}} \\
  \quad\quad\quad\case{\infer{ }{A\leq A}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
  \quad\quad\quad\case{\_}
  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
      : A_1}}{\vdash \app M N : B}}
\end{mathleft}

\noindent
The $\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\
\prd A {\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
construct a subtyping derivation for its argument, or failing with an
error (definition not included here). Notice the slight optimization
to the size of the derivation: if the types $A_1$ and $A_2$ are the
same syntactically, we don't generate a subtyping rule. This is done by
matching on the returned derivation $\md_\leq$.

@@

For what concerns the treatment of environments, the case of $\recb M
N x y P$ is similar: we first compute recursively the derivations
$\md_M$ and $\md_N$ which should be well-typed in the current
environment, and then $\md_P$ in the enlarged environment, with
variables $\var x$ and $\var y$ in $P$ substituted with the (future)
result of $\finfer{}$. The problem is with subtyping: terms $N : A_N$
and $P : A_P$ should be coercible to the same type $A$, but $P$ should
be typed under $\vdash y:A$, which is not known before typing $P$. The
trick to break the loop is to iterate the typing of $P$ with more and
more general types, until we reach a fixpoint (this is finite if there
is a unique more general type); in our case, that means iterating
twice: we will then attain \cst{nat}.

\begin{mathleft}
  \quad\case{\recb M N x y P} \\
  \quad\quad
  \letd {\pair {A_M} {\md_M}} {\finfer{M}} \\
  \quad\quad\letd {\md_{A_M}} {{A_M} \leq \cst{nat}} \\
  \quad\quad
  \letd {\pair {A_N} {\md_N}} {\finfer{N}} \\
  \quad\quad
  \syntax{let} {\pair {A_P} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_N)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A_N}{\md_y}}}}}} \syntax{in}\\
  \quad\quad\letd {\pair A {\pair{\md_{A_N}}{\md_{A_P}}}} {{A_N} \sqcap {A_P}} \\
  \quad\quad
  \syntax{let} {\pair {\_} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A}{\md_y}}}}}} \syntax{in}\\
  \hspace{-2em}\pair {A} {
    \infer{
      \infer*{\md_M \and \md_{A_M}}
      {\vdash M : A} \and
      \infer*{\md_N \and \md_{A_N}}
      {\vdash N : A}\and
      \infer*{\infer*{}{\mbox{$
        \begin{array}{c}
          [\md_x] [\md_y] \\
          \md_P
      \end{array}
        $}} \and \md_{A_P}}{\vdash P : A}
    }{
      \vdash \recb M N x y P : A
    }
  }
\end{mathleft}

\noindent
Again, we use a new auxiliary function
$$\sqcap\ :\ \prd{A}{\cst{tp}} \prd{B}{\cst{tp}} \sig C {\cst{tp}}
(\vdash A\leq C )\times(\vdash B\leq C) = \ldots$$ computing the sup
of two types $A$ and $B$, and returning proofs of of subtyping. Note
that we didn't perform here the optimization seen in the application
case.

@@

We exploit here an idea similar to run-time
\emph{contracts}: %TODO cite
there is no static guarantee that computations won't return ill-typed
values, but arguments and results are checked at run-time at the
boundaries of defined functions. This allows to omit formal
justifications, like the fact that we need to iterate twice the typing
of $P$ in $\recb M N x y P$, or that $B$ can't depend on $\md_x$ in
the $\lambda$ case. Of course this incurs on the performance; we will
see a compensation for it by the slicing mechanism described below in
\ref{sec:repr}.

A similar idea is very wide-spread in \sysname{LCF}-style proof
assistants: during manual proof search, the interaction is driven by
\emph{tactics}: %TODO cite
they are untrusted programs generating pieces of proofs, assembled by
other, higher-order tactics; verification of the generated proofs is
done when the evaluation of all tactics is complete. The differences
with our mechanism are two-fold: first, tactics usually take one (or
several) \emph{goal}(s) as input, \ie\ holes in a proof, whereas our
functions can take any term; secondly, proof-check is usually done at
the end of the evaluation, which can lead to less tractable bugs; by
making evaluation and checking mutually recursive, we choose to detect
errors earlier in these witnesses: exactly at the point where they are
generated.

Introducing the ``inverse'' of the derivation inference function has
two purpose: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would arrive to the unhandled case $\finfer{\var
  x}$. This requires evaluating the type-checker with a carefully
designed strategy (see \ref{sec:comput}).

% TODO cite lee2007mechanizing

@@

\subsection{The computational calculus}
\label{sec:comput}
